---
title: "Spatial interaction models with R"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Spatial interaction models with R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Prerequisites

You need to have [installed R](https://cran.r-project.org/) and a suitable editor such as [RStudio](https://www.rstudio.com/products/rstudio/download/preview/) or [VSCode with R plugin](https://github.com/REditorSupport/vscode-R).
See the package's [README](https://robinlovelace.github.io/si/) for instructions on installing the {si} package.

<!--#  Test the necessary packages are installed by loading them (we will also use {dplyr}):-->

```{r setup}
library(si)
```

# Input data

This tutorial builds on a [reproducible](https://github.com/adamdennett/SIModelling/blob/SIModelling-Edits/SimAus2.Rmd) guide to SIMs in R [@dennett_modelling_2018].^[
See @dennett_estimating_2012 or an earlier guide on SIMs in R.
]
We start by importing open access data representing movement between zones in Australia (thanks to Adam Dennett for making the files accessible):

```{r import}
u1 = "https://github.com/Robinlovelace/si/releases/download/0.0.1/zones_aus.geojson"
zones_aus = sf::read_sf(u1)
u2 = "https://www.dropbox.com/s/wi3zxlq5pff1yda/AusMig2011.csv?raw=1"
od_aus = read.csv(u2)
```

```{r, echo=FALSE, eval=FALSE}
u1 = "https://www.dropbox.com/s/0fg80nzcxcsybii/GCCSA_2016_AUST_New.geojson?raw=1" # original
zones_original = sf::read_sf(u1)
sf::write_sf(zones_original, "GCCSA_2016_AUST_New.geojson")
lobstr::obj_size(zones_original$geometry) / lobstr::obj_size(zones_original)
# 99% of size taken up with geometry, no need for spatial resolution -> simplify
piggyback::pb_upload("GCCSA_2016_AUST_New.geojson")
readr::write_csv(od_aus, "od_migration_original.csv")
piggyback::pb_upload("od_migration_original.csv")
library(sf)
zones_aus_simplified = rmapshaper::ms_simplify(zones_original, keep = 0.001)
lobstr::obj_size(zones_aus_simplified) / lobstr::obj_size(zones_original)
# 0.2% of the size
plot(zones_aus_simplified$geometry[1]) # still too detailed?
sf::write_sf(zones_aus_simplified, "zones_aus.geojson")
piggyback::pb_upload("zones_aus.geojson")
```

Let's take a quick look at and 'minimize' these input datasets before modelling them with SIMs:

```{r clean}
dim(zones_aus)
names(zones_aus)
key_zone_names = c("GCCSA_CODE", "GCCSA_NAME", "AREA_SQKM")
zones = zones_aus[key_zone_names]
head(zones, 2)
dim(od_aus)
names(od_aus)
key_od_names = c("Orig_code", "Dest_code", "Flow")
od = od_aus[key_od_names]
head(od, 2)
```

The results printed above show that:

-   `zones_aus` represents the 15 regions of Australia, with columns containing zone identifiers (IDs), primarily `GCCSA_CODE`, and the area of each zone (`AREA_SQKM`). The minimal `zones` object contains only the region code, name, and area.

-   `od_aus` contains 225 rows of data representing the movement of people between the zones in `aus`.
    Note that 255 is 15 squared, meaning that this OD dataset contains the complete combination of migration flows, starting with zone `1GSYD` to `1GSYD`.
    These codes are present in the `GCCSA_CODE` column in the `aus` object.
    The first row of the OD dataset represents 'intra-zonal' migrations in Greater Sydney, presumably counting the number of people who move house from somewhere in the region to another home in the region.
    There are 13 columns in this dataset, the most important of which are `Orig_code`, `Dest_code`, and `Flow`, which we have captured in the minimal `od` dataset.

Note: a useful convention with 'long form' OD datasets is for the first two columns to contain zone IDs that correspond to values in the first column of the zone dataset.
The R package [{od}](https://itsleeds.github.io/od/), on which {si} builds, assumes that inputs to its functions are in this form.

It is a good idea to verify that the origin and destination codes in the `od` dataset match the zone codes in `zones`:

```{r}
summary(od[[1]] %in% zones[[1]])
summary(od[[2]] %in% zones[[1]])
```

It is clear from the above that we have 'clean' input datasets, let's begin with the modelling!

# Preparing a SIM

Key to the workings of the {si} package is the conversion of geographic objects representing origins and destinations into an OD dataset.
In this case, we already have an OD dataset, so this step is less relevant.
However, we will take this step in any case because many SIMs start without such a comprehensive OD dataset as we have in this case.

Prepare the OD dataset as follows:

```{r}
od_sim = si_to_od(origins = zones, destinations = zones)
```


# References
