---
title: "Tutorial: spatial interaction models with R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: spatial interaction models with R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Pre-requisites: you should have [installed R](https://cran.r-project.org/) and a suitable editor such as [RStudio](https://www.rstudio.com/products/rstudio/download/preview/) or [VSCode with R plugin](https://github.com/REditorSupport/vscode-R).
See the package's [README](https://robinlovelace.github.io/si/) for instructions on installing the {si} package.

```{r setup}
library(si)
```

This tutorial builds on @dennett_modelling_2018 (which in turn builds on @dennett_estimating_2012's pioneering [reproducible](https://github.com/adamdennett/SIModelling/blob/SIModelling-Edits/SimAus2.Rmd) tutorial on SIMs with R).
We start by importing open access data representing movement between zones in Australia (thanks to Adam Dennett for making the files accessible):

```{r}
u1 = "https://www.dropbox.com/s/0fg80nzcxcsybii/GCCSA_2016_AUST_New.geojson?raw=1"
aus = sf::read_sf(u1)
u2 = "https://www.dropbox.com/s/wi3zxlq5pff1yda/AusMig2011.csv?raw=1"
od_migration_original = read.csv(u2)
```

```{r, echo=FALSE, eval=FALSE}
sf::write_sf(aus, "GCCSA_2016_AUST_New.geojson")
piggyback::pb_upload("GCCSA_2016_AUST_New.geojson")
```
