---
title: "Validation of spatial interaction models"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(si)
library(sf)
library(tmap)
library(dplyr)
```

# Validation of 2011 OD data for Leeds, UK

The input datasets for this example are as follows:

- Geographic boundary and centroid data from the [{pct}](https://itsleeds.github.io/pct/) package, trip representing origins and destinations
- Origin-destination data representing trips to work from the 2011 Census
- Explanatory variables associated with zones, tbc

The key input datasets are shown below

```{r}
si_od_census
plot(si_zones$geometry)
plot(si_centroids$geometry, add = TRUE)
```

```{r}
od_from_si = si_to_od(origins = si_centroids, destinations = si_centroids)
nrow(si_od_census) / nrow(od_from_si)
```

The output shows that more than 90% of all possible combinations of OD pairs going from and to every zone are present in the validation dataset.
The maximum number of unique OD pairs for a given OD dataset is the number of origins multiplied by the number of destinations.
In this case that is `r nrow(si_zones)` * `r nrow(si_zones)` which is `r nrow(si_zones)^2` 
Before proceeding it's worth doing some sanity checks to find out:

- What does the spatial distribution of travel to work in Leeds look like?
- Which OD pairs that are 

```{r}
si_od_census_sf = si_od_census %>% 
  od::od_to_sf(si_centroids) %>%
  mutate(all, `Percent active` = (bicycle + foot) / all * 100)

si_od_census_sf %>%
  arrange(`Percent active`) %>% 
  tm_shape() +
  tm_lines(col = "Percent active", lwd = "all", scale = 3, palette = "viridis",
           breaks = c(0, 0.1, 0.2, 0.4, 1) * 100, legend.lwd.show = FALSE) +
  tm_scale_bar()
```

The figure above clearly shows that Leeds is a monocentric city with relatively high levels of active travel near the centre but with low levels of active travel in the outskirts.
From that we can predict the ~8% of OD pairs that *are not* in the dataset: long distance OD pairs avoiding the centre.
Let's find out:

```{r}
od_pairs_with_no_travel = od_from_si %>% 
  filter(!paste(O, D) %in% paste(si_od_census$O, si_od_census$D))
summary(od_pairs_with_no_travel$distance_euclidean)
summary(sf::st_length(si_od_census_sf))
```

As expected, the mean length of OD pairs in the observed Census is substantially less, and that's before calculating the weighted mean, which is just over 5 km:

```{r}
weighted.mean(sf::st_length(si_od_census_sf), w = si_od_census_sf$all)
```

In any case, let's proceed to some estimates of travel.




# Valitation based on 2011 OA-WPZ data
